---
title: "Computing inter-cellular distances"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: kable
    toc: true
vignette: >
  %\VignetteIndexEntry{Computing inter-cellular distances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo=FALSE,include=FALSE,message=FALSE}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=7, fig.height=5, 
                      comment=NA, warning=FALSE, message=FALSE)
theme_set(theme_bw())
# No figure margins
par(mar=rep(0, 4))
```

<style type="text/css">
img { 
  border: none;
}
table {
    width: auto !important;
}
</style>

`phenoptr` contains several functions which analyze and report on the spatial
relationship between cells in a single field.

## Nearest neighbor distances

### Computing with `find_nearest_distance`

The `find_nearest_distance` function finds per-cell nearest neighbor distances.
For each cell in a sample, it finds the distances to the nearest neighbor
cells in each of the provided phenotypes. 
The distances are returned in a `data_frame` with
one column per phenotype.

For example, the `phenoptr` sample data `sample_cell_seg_data` contains
`r length(unique(phenoptr::sample_cell_seg_data$Phenotype))` unique
phenotypes:

```{r}
library(tidyverse)
library(phenoptr)

csd <- sample_cell_seg_data
csd %>% count(Phenotype)
```

The `other` cells are not of 
interest, so first filter them out.

```{r}
csd <- csd %>% filter(Phenotype!='other')
```

Calling `find_nearest_distance` on this file returns a `data_frame` with 
four columns and one row for each cell:

```{r}
distances <- find_nearest_distance(csd)
glimpse(distances)
nrow(csd)
```

To create a combined data frame, use `dplyr::bind_cols`.

```{r}
csd_with_distance <- bind_cols(csd, distances)
```

Note: Cell positions in a cell seg data file are relative to the 
top-left corner of the field, 
so calling `find_nearest_distance` on a merged data file will 
compute incorrect values. One 
way to add distance columns to a merged data file is to use `dplyr::group_by`
to process each field separately. For example, if `merged` contains merged
cell seg data, add distance columns with this code:

```{r eval=FALSE}
merged_with_distance <- merged %>%
  dplyr::group_by(`Sample Name`) %>%
  dplyr::do(dplyr::bind_cols(., find_nearest_distance(.)))
```

### Analyzing nearest neighbor distances

Once the nearest neighbors have been computed per cell, standard aggregation,
analysis and plotting commands can be used to examine the results. For example,
find the mean nearest neighbor distances by phenotype:
 
```{r}
csd_with_distance %>% group_by(Phenotype) %>% 
  select(starts_with('Distance to')) %>% 
  summarize_all(~round(mean(.), 1))
```

Show the distribution of distances in a density plot:

```{r}
ggplot(csd_with_distance, aes(`Distance to CD8+`, color=Phenotype)) +
  geom_density(size=1) + theme_minimal()
```

----

## Cells within a radius

### Computing with `count_within`

The `count_within` function looks at the number of cells within a radius of
another cell and returns summary measures. For example, use `count_within`
to find the number of `CD68+` cells
having a `CK+` cell within 25 microns:

```{r}
count_within(csd, from='CD68+', to='CK+', radius=25)
```

In this result, `from_count` and `to_count` are the total numbers of 
eligible cells. They agree with the counts in the first table in this
tutorial. `from_with` is the number of `CD68+` cells having at least 
one `CK+` cell within 25 micron. `within_mean` is the average number of
`CK+` cells found within 25 micron of each `CD68+` cell.

Note there are some subtleties to `count_within`. Most importantly, it
is not symmetric. In this example, the number of `CK+` cells with a 
`CD68+`
within 25 microns is not the same as the number of `CD68+`
cells with a `CK+` cell within 25 microns.

```{r}
count_within(csd, from='CK+', to='CD68+', radius=25)
```

Please see `help(count_within)` for details.

### Count multiple files or phenotypes using `count_within_batch`

You may want to run `count_within` on an entire directory of cell seg data
files, or to count multiple combinations of phenotypes. Both of
these are possible using `count_within_batch`. This function takes 
the path to a directory that contains multiple cell seg data files.
The `pairs` and `category` parameters are lists and may contain
multiple entries.

For example, the following commands will count `FoxP3+` cells
with a `CD8+` or `CK+` cell within 10 or 25 microns.
Separate counts are returned for each `to` phenotype and for
`Tumor` and `Stroma` tissue categories. Counts will be calculated 
for all `cell_seg_data.txt` files in `my_directory`.

```{r eval=FALSE}
base_path <- "/path/to/my_directory"

pairs <- list(c('FoxP3+', 'CD8+'),
              c('FoxP3+', 'CK+'))
radii <- c(10, 25)
categories <- c('Stroma', 'Tumor')

count_within_batch(base_path, pairs, radii, categories)
```

See `help(count_within_batch)` and the tutorial
[Selecting cells within a cell segmentation table](https://perkinelmer.github.io/phenoptr/articles/selecting_cells.html)
for additional details.

----

## Visualizing nearest neighbors

The `spatial_distribution_report` function is a bit different from the other
functions
mentioned in this tutorial Rather than calculate and return distance metrics,
it creates a report which shows visually the nearest neighbor relations between
two phenotypes in a single field. Because the result is a stand-alone HTML file,
it can't easily be demonstrated in a tutorial For an example, 
copy and paste this code
into your own copy of R. It will create a sample report in your user directory.

```{r eval=FALSE}
cell_seg_path <- sample_cell_seg_path()

pairs <- list(c('CD68+', 'CD8+'))
colors <- c('CD68+'='magenta', 'CD8+'='yellow')
out_path <- path.expand('~/spatial_distribution_report.html')

spatial_distribution_report(cell_seg_path, pairs, colors, output_path=out_path)

```

To create reports for all cell seg data files in a directory, first define
`phenotypes` and `colors` as above. Use `list_cell_seg_files` to find all
the files. Then call `spatial_distribution_report` for each file. This will
create reports in the same directory as the data files.

```{r eval=FALSE}
base_path <- '/path/to/data/'
paths <- list_cell_seg_files(base_path)
for (path in paths)
  spatial_distribution_report(path, pairs, colors)
```

