---
title: "Find and count touching cells"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: kable
    toc: true
vignette: >
  %\VignetteIndexEntry{Find and count touching cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,echo=FALSE}
knitr::opts_chunk$set(eval=FALSE)
```


The `count_touching_cells` function uses morphological analysis of nuclear and
membrane segmentation maps to find touching cells of paired phenotypes. It
reports the number of touching cells found and, optionally, writes image files
showing the touching cells.

`count_touching_cells` uses the results of inForm cell segmentation to
determine which cells are touching. It uses both nuclear and membrane
segmentation to determine the extent of each cell.

`count_touching_cells` processes a single field and multiple pairs of 
phenotypes. The specification of pairs and phenotypes is flexible to 
accommodate any requirement. The simplest case uses the phenotype names
from inForm to select cells. For this case, only the `pairs` argument
is needed. 

For example, this code finds and visualizes touches between 
`tumor` and `CD8` cells, and also
between `tumor` and `CD68` cells:

```{r pairs_only}
cell_seg_path = '/path/to/cell_seg_data/'
pairs = list(
  c('tumor', 'CD8'),
  c('tumor', 'CD68')
)
colors=list(tumor='cyan', CD8='yellow', CD68='magenta')
count_touching_cells(cell_seg_path, pairs, colors)
```

For more flexibility, use the `phenotype_rules` argument. For example,
this code repeats the previous analysis limiting it to tumor cells with PDL1
above a threshold. Note that `phenotype_rules` only needs to include
definitions for phenotypes which don't match the names in `pairs`.
(See `vignette('selecting_cells', package='phenoptr')` for more details on
selecting pairs.)

```{r phenotype_rules}
pairs = list(
  c('tumor PLD1+', 'CD8'),
  c('tumor PLD1+', 'CD68')
)

phenotype_rules = list(
  'tumor PLD1+'=list('tumor', ~`Entire Cell PDL1 (Opal 620) Mean`>0.5)
)
colors=list('tumor PLD1+'='cyan', CD8='yellow', CD68='magenta')
count_touching_cells(cell_seg_path, pairs, colors, phenotype_rules)
```

Using `purrr::map`, you can find touching cells for all cell seg data
files in a single directory.

```{r process_directory}
# Directory containing data files
base_path = '/path/to/data'

# A subdirectory for the results
output_base = file.path(base_path, 'touches')

# All cell seg data files in base_path
files = list_cell_seg_files(base_path)

# Count and visualize touching cells
touch_counts = purrr::map_df(files, function(path) {
  cat('Processing', path, '\n')
  count_touching_cells(path, pairs, colors, phenotype_rules,
                       output_base=output_base)
})
```

The result of the above is a `data_frame` which may be written to 
a CSV file:

```{r write_csv}
touches_path = file.path(output_base, 'TouchCounts.csv')
readr::write_csv(touch_counts, touches_path)
```
