---
title: "Reading and displaying inForm image Files"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    fig_caption: yes
    toc: true
vignette: >
  %\VignetteIndexEntry{Reading and displaying inForm image Files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE,include=FALSE,message=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5, 
                      comment=NA, warning=FALSE, message=FALSE)
# No figure margins
par(mar=rep(0, 4))
```

<style type="text/css">
img { 
  border: none;
}
</style>

inForm writes a variety of image files in JPEG and TIFF format. This
tutorial shows how to read the image files in R.

The R packages `tiff` and `jpeg` provide functions `readTIFF` and `readJPEG`
which read image data into R matrices. `phenoptr` expands on these with
`read_components` and `read_maps` which read multiple images and extract
useful metadata from the Image Description fields.

A variety of functions can display image data including `plot.raster`,
`ggplot2::annotation_raster` and `EBImage::display`. 

----

## Coordinate systems

A few items to note:

- Images have the origin (0, 0) at the top left. This is consistent with
the `Cell X Position` and `Cell Y Position` columns of an inForm
cell seg table. The _row_ (first) dimension of an array corresponds to
the $y$ axis and `Cell Y Position` column. The _column_ (second) array dimension
corresponds to the $x$ axis and the `Cell X Position` column. 

- Image dimensions are in pixels. Cell seg tables are written with pixel 
dimensions and converted to microns when read by `phenoptr::read_cell_seg_data`.

- Image display methods differ in their orientation; some show $x$ and $y$ as
described above; others transpose the image.

Each of these items easily causes confusion; be careful and verify your work!

----

## Reading and displaying color images

Color (RGB) images output from inForm, such as the composite image and 
segmentation images, can be read directly using 
`tiff::readTIFF()` or `jpeg::readJPEG()`.
For example,

```{r read_composite}
path <- system.file("extdata", "sample", 
                   "Set4_1-6plex_[16142,55840]_composite_image.jpg", 
                   package = "phenoptr")
img <- jpeg::readJPEG(path)
dim(img)
```

Color images are read as 3D arrays. They can be displayed by converting to
a raster. 

```{r show_composite}
snippet <- as.raster(img[800:1000, 1200:1500,])
plot(snippet)
```

`EBImage::display` will display an array in a web browser. RGB images are 
shown as three grayscale planes unless converted to an `EBImage::Image` first.
`EBImage::display` shows the first coordinate horizontally
and the second coordinate vertically so the image must be flipped to display
in the expected orientation.

```{r show_composite_ebimage,eval=FALSE}
img_transposed <- aperm(img, c(2, 1, 3))
EBImage::display(img_transposed)
EBImage::display(EBImage::Image(img_transposed, colormode='Color'))
```

----

## Reading and displaying component data

inForm saves component data as multiple 32-bit floating-point
images within a single TIFF file.
Images smaller than 2K by 2K pixels are saved in "strip" format and may
be read by `tiff::readTIFF()`. Larger images are saved in "tiled" format
which is not supported by `readTIFF()`. Single fields from Vectra Polaris,
Vectra 3 and Mantra are all smaller than 2K by 2K and may be read
by these functions; larger fields may not read.  Install the 
[Akoya Biosciences fork](https://github.com/akoyabio/tiff) of the tiff 
package to support tiled images and remove this limitation.

`phenoptr::read_components()` is a wrapper around `readTIFF()` which reads
the individual component images from a `component_data.tif` file. 
It keeps the full-resolution component images,
extracts the component names from the image descriptions, 
and returns a named list of image matrices.

Component data may be displayed using `plot.raster` or `EBImage::display`. 
See the next section for examples of displaying grayscale images such as these.

----

## Reading and displaying segmentation maps

inForm saves segmentation maps (nuclear segmentation, etc.) as multiple
16-bit grayscale images within a single TIFF file. Most segmentation images
are _label_ images, where each object is represented by a region whose value
is the object number. The membrane map is a binary image where the presence
of membrane is indicated with a "1" value.

`phenoptr::read_maps()` is a wrapper around `readTIFF()` which reads
map files. It returns a named list of integer-valued matrices. The
list names reflect the content of the individual images, e.g. Nucleus,
Cytoplasm, etc.

```{r read_map}
map_path <- system.file("extdata", "sample",
   "Set4_1-6plex_[16142,55840]_binary_seg_maps.tif", package = "phenoptr")
maps <- phenoptr::read_maps(map_path)
names(maps)
```

The Nucleus label image contains values for background (0) and each Cell ID.

```{r nucleus}
nucleus <- maps[['Nucleus']]
dim(nucleus)
range(nucleus)
range(phenoptr::sample_cell_seg_data$`Cell ID`)
```

Label images may be displayed with `plot.raster` or `EBImage::display`
(which requires swapping axes).


```{r display_map}
nucleus_snippet <- nucleus[800:1000, 1200:1500]
plot(as.raster(nucleus_snippet, max=max(nucleus_snippet)))
```

```{r display_map_ebimage,eval=FALSE}
EBImage::display(t(nucleus/max(nucleus)))
EBImage::display(t(maps[['Membrane']]))
```

----

## Plotting image data

Images have the origin (0, 0) at the top left. This is consistent with
the `Cell X Position` and `Cell Y Position` columns of an inForm
cell seg table but it is reversed (in $y$) from the usual plotting conventions.
If you plot images and points together, you will have to reverse one or
the other. 

Here is an example using `ggplot2`. The $y$-axis is reversed
using `ggplot2::scale_y_reverse()` and the image is then un-reversed
by negating its $y$ limits.

```{r plotting}
library(ggplot2)
csd <- phenoptr::sample_cell_seg_data
csd <- subset(csd, `Cell X Position`>=600 & `Cell X Position`<=750 &
              `Cell Y Position`>=400 & `Cell Y Position`<=500)
ggplot(data=csd, aes(`Cell X Position`, `Cell Y Position`, color=Phenotype)) +
  scale_x_continuous(limits=c(600, 750)) + 
  scale_y_reverse(limits=c(500, 400)) +
  annotation_raster(snippet, 600, 750, -400, -500) +
  geom_point(size=2) + coord_equal() +
  scale_color_manual(values=c("CK+"="cyan", "CD8+"="yellow",
                              "other"="blue", "CD68+"="magenta", 
                              "FoxP3+"="orange")) +
  theme_minimal()
```

