---
title: "Reading and exploring inForm tables"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    df_print: kable
    toc: true
vignette: >
  %\VignetteIndexEntry{Reading and exploring inForm tables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, echo=FALSE,include=FALSE,message=FALSE}
suppressPackageStartupMessages(library(tidyverse))
knitr::opts_chunk$set(fig.width=6, fig.height=4, 
                      comment=NA, warning=FALSE, message=FALSE)
theme_set(theme_bw())
# No figure margins
par(mar=rep(0, 4))
```

<style type="text/css">
img { 
  border: none;
}
table {
    width: auto !important;
}
.fa-external-link {
    font-size: 10px;
}
</style>

## inForm tables

PerkinElmer's inForm<sup>&reg;</sup> software exports a variety of
tabular data including information about cells and tissue categories.
Tables are exported as tab-delimited text files with file names
derived from the name of the source image by adding a suffix. Some
of the suffixes, and the data contained in the files, are

- `cell_seg_data.txt` - Detailed information about each cell in a field,
including location, size, expression data for each component and possibly
the phenotype and the containing tissue category.
- `cell_seg_data_summary.txt` - Summary information about the cells in a 
field, including cell counts (stratified by phenotype and tissue category,
if available) and summary size and expression data.
- `tissue_seg_data.txt` - Detailed information about each tissue category
region in the field, including size, centroid and expression data.
- `tissue_seg_data_summary.txt` - Summary information about tissue category
regions.

## Cell segmentation data

`phenoptr` is primarily concerned with loading and processing 
`cell_seg_data.txt` files created by inForm. In the package documentation,
these files are called _cell seg data files_ and the data within them is referred
to as _cell seg data_. Most examples use a variable named `csd` to 
contain cell seg data.

### Reading cell segmentation data files

The `read_cell_seg_data` function reads cell seg data files and does
useful cleanup on the result including removing empty columns,
converting pixels to microns and simplifying column names.

```{r read}
# Load libraries
library(tidyverse)
library(phenoptr)

# sample_cell_seg_path gives the path to a sample file included with phenoptr.
# Change this to be the path to your data.
path <-  sample_cell_seg_path()

# Read the data file
csd <- read_cell_seg_data(path)

# Show some nicely shortened names
# The suffix "(Normalized Counts, Total Weighting)" has been removed.
grep('Nucleus.*Mean', names(csd), value=TRUE)
```

The value returned from `read_cell_seg_data` is a `data_frame` (also known
as a `tibble` 
[<i class='fa fa-external-link'></i>](http://tibble.tidyverse.org/)
- the
[tidyverse](http://tidyverse.org/) version of a `data.frame`) and can be
inspected as you would any other `data_frame`.

```{r inspect}
# How many cells did we read?
nrow(csd)

# How many cells of each phenotype are in each tissue category?
table(csd$`Tissue Category`, csd$Phenotype)
```

Notice that column names containing spaces or other special characters,
such as `Tissue Category` in the example above, generally must be enclosed in
backticks (`) in code. If you are editing in
[RStudio](https://www.rstudio.com/products/rstudio/) (highly recommended),
tab-completion will often include the backticks for you.

### Reading cell segmentation summary files

`read_cell_seg_data` is also useful for reading `cell_seg_data_summary.txt`
files. The same cleanup done for cell seg files is helpful for summary files.

## Creating new columns

The `dplyr::mutate`
[<i class='fa fa-external-link'></i>](http://dplyr.tidyverse.org/reference/mutate.html)
function makes it very easy to add new columns to your
data. This example adds a column for PDL1 positivity and counts cells
in each phenotype stratified by positivity.

```{r mutate}
csd <- csd %>% mutate(pdl1_plus=`Entire Cell PDL1 (Opal 520) Mean`>3)
table(csd$pdl1_plus, csd$Phenotype)
```

### The pipe operator `%>%`

These examples make extensive use of the pipe operator (`%>%`) to combine
operations. If you are not familiar with this operator, you may
want to read this 
[introduction](http://magrittr.tidyverse.org/#usage).

## Aggregating data

The functions `dplyr::group_by`
[<i class='fa fa-external-link'></i>](http://dplyr.tidyverse.org/reference/group_by.html)
and `dplyr::summarize`
[<i class='fa fa-external-link'></i>](http://dplyr.tidyverse.org/reference/summarise.html)
can be used to aggregate
data within groups. `dplyr::filter`
[<i class='fa fa-external-link'></i>](http://dplyr.tidyverse.org/reference/filter.html)
removes unwanted values.
This example computes the mean PDL1 expression for each
phenotype, omitting `other` cells.

```{r aggregate}
csd %>% 
  filter(Phenotype!='other') %>% 
  group_by(Phenotype) %>% 
  summarize(mean_pdl1=mean(`Entire Cell PDL1 (Opal 520) Mean`))
```

## Plotting data

[ggplot2](http://ggplot2.tidyverse.org/) is a powerful and versatile 
plotting package well suited to exploratory data visualization. Here are
two brief examples to whet your appetite.

```{r ggplot2}
ggplot(csd, aes(Phenotype, `Entire Cell PDL1 (Opal 520) Mean`, color=Phenotype)) +
  geom_boxplot() + 
  scale_color_brewer(palette='Set1') + 
  labs(y='PDL1 Expression', title='PDL1 Expression per Phenotype')

ggplot(csd %>% filter(Phenotype!='other'), 
       aes(`Entire Cell PDL1 (Opal 520) Mean`, 
                `Entire Cell PD1 (Opal 650) Mean`,
                color=Phenotype)) +
  geom_point(size=1, alpha=0.2) + 
  facet_wrap(~Phenotype) +
  scale_x_log10() + scale_y_log10() + scale_color_brewer(palette='Set1') +
  labs(x='PDL1 Expression', y='PD1 Expression', 
       title='Comparison of PD1 and PDL1 Expression per Phenotype')
```

## Further reading and examples

The examples here only scratch
the surface of what `dplyr`, `ggplot2` and other functions in the `tidyverse`
can do. If you'd like to learn more about the tidyverse, 
a good place to start is Garrett Grolemund and Hadley Wickham's book,
available free online at
[R for data science](http://r4ds.had.co.nz/). 
The [Data transformation](http://r4ds.had.co.nz/transform.html) chapter
introduces the `dplyr` functions used in this tutorial.

The [Tutorials](https://perkinelmer.github.io/phenoptrExamples/index.html)
included with the `phenoptrExamples` package include examples of
reading data from multiple fields and
aggregating across fields. The package includes more extensive 
sample data which can be used for practice.
