---
title: "Selecting cells within a cell segmentation table"
author: "Kent Johnson"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Selecting cells within a cell segmentation table}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup,echo=FALSE}
#knitr::opts_chunk$set(eval=FALSE)
```

`phenoptr` includes a flexible mechanism for selecting cells (i.e. rows) from a
cell seg table. The mechanism is implemented in `select_rows`. Row selection
may be used directly via `select_rows` and ordinary subsetting operations. It is 
also used indirectly by calling functions that support it, 
including `count_touching_cells` and `count_within`.

The return value from `select_rows` is a boolean (logical) vector whose 
length is the number of rows of the given cell seg table. 
You use this returned
value to select rows of the table. 

<div class="panel panel-default"><div class="panel-body">
The mechanism `select_rows` uses to specify phenotypes is very flexible. This
flexibility comes at a cost in complexity. Most common phenotype combinations
can also be specified using `parse_phenotypes`. 
</div></div>

----

## Selecting phenotypes

This tutorial uses `sample_cell_seg_data` and 
`count_within` to give examples of the phenotype specifications used by 
`select_rows`.

### Select a single phenotype

The simplest selector is just the name of a single phenotype. This example
selects the rows containing `CK+` cells. The same syntax works with 
`parse_phenotypes`.

```{r}
library(phenoptr)

csd <- sample_cell_seg_data
rows <- select_rows(csd, 'CK+')
sum(rows) # The number of selected rows

# Select just the desired rows by subsetting
ck <- csd[rows, ]
dim(ck)
```

This example counts `CD8+` cells with 15 microns of `CK+` cells.

```{r}
dst <- distance_matrix(csd) # Compute this just once and re-use it
count_within(csd, from='CK+', to='CD8+', radius=15, dst=dst)
```

### Select multiple positivity

Double positive (or more) cells can be selected by including multiple
names in a list. Selectors in a list are combined with AND.

### Group multiple phenotypes

Multiple phenotypes may selected together by including each name in a 
character vector (not a list!). Names in a vector are combined with OR.

For example, to select cells phenotyped as either 
`CD8+` or `FoxP3+`,
use the selector `c('CD8+', 'FoxP3+')`.

This example selects this combination. Note the call to `select_rows`
has been combined with the subsetting of `csd`.

```{r}
tcells <- csd[select_rows(csd, c('CD8+', 'FoxP3+')), ]
dim(tcells)

count_within(csd, from='CK+', to=c('CD8+', 'FoxP3+'), radius=15, dst=dst)
```

This type of grouping is an either / or selection. The `count_within` 
example above counts the number of T cells (`CD8+` _or_ `FoxP3+`) 
within 15 microns of a `CK+`
cell. If you want separate counts for `CD8+` and `FoxP3+`, use
`count_within_batch`.

### Flexible selection using expressions

For more flexibility, `select_rows` supports selection using any valid
R expression. Expressions are written using one-sided formulas. 
The formulas are evaluated in the context
of the cell seg table so they may reference any column of the table.

For example, to select cells with PDL1 expression greater than 3, use
the expression ``~`Entire Cell PDL1 (Opal 520) Mean`>3``. 
In this example, the column name is `Entire Cell PDL1 (Opal 520) Mean`.

Expressions and phenotype names may be combined in a list. This example
selects `CK+` cells with PDL1 > 3.

```{r}
rows <- select_rows(csd, list('CK+', ~`Entire Cell PDL1 (Opal 520) Mean`>3))
ck_pdl1 <- csd[rows, ]
dim(ck_pdl1)

count_within(csd, from=list('CK+', ~`Entire Cell PDL1 (Opal 520) Mean`>3), 
             to='CD8+', radius=15, dst=dst)
```

A few things to note about formula expressions:

- Expressions are evaluated in the context of the cell seg table. Names
  used in the expression must match column names in the table. If the table
  was read using `read_cell_seg_data(path, remove_units=TRUE)` (the default),
  the table names will be abbreviated compared to the names in the file.
- Names which are not valid R symbol names---this includes most of the column
  names in a cell seg table---must be enclosed in backticks (`) as
  in the example above.
  
----

## Selecting pairs of phenotypes

Several functions in `phenoptr` operate on pairs of phenotypes 
and have arguments
`pairs` and `phenotype_rules`. For example, see `count_touching_cells`
and `spatial_distribution_report`.
These functions build on `select_rows` to
allow allow flexible selection of pairs of phenotypes. 

### Pairs of existing phenotypes

In the simplest usage, the names in `pairs` are the names of phenotypes in
the cell seg data. In this case, `pairs` just lists the desired phenotypes.
For example, to pair `CK+` cells with `CD8+` cells, use the argument

```{r}
pairs <- list(c('CK+', 'CD8+'))
```

For a single pair, a list is not required so this can be simplified to

```{r}
pairs <- c('CK+', 'CD8+')
```

For multiple pairs, list each pair separately. For example, to pair `CK+` 
cells first with `CD8+` cells and then with `CD68+` cells, use the argument

```{r}
pairs <- list(c('CK+', 'CD8+'),
             c('CK+', 'CD68+'))
```

### Defining new phenotypes

You may want to define a new phenotype using grouping or expressions as shown 
in the "Selecting phenotypes" sections above. To do
this, use the `phenotype_rules` argument to associate a `select_rows` rule
with a name; then use the new name in the pairs argument.

For example, to create a `T Cell` phenotype which matches `CD8+` and `FoxP3+`
phenotypes, and pair it with a `PDL1+ CK+` phenotype which 
applies a threshold to
tumor cells, use these arguments:

```{r}
pairs <- c('PDL1+ CK+', 'T Cell')
phenotype_rules <- list(
  'PDL1+ tumor'=list('CK+', ~`Entire Cell PDL1 (Opal 520) Mean`>3),
  'T Cell'=c('CD8+', 'FoxP3+'))
```

`phenotype_rules` only needs to include phenotypes which are not in the cell
seg data. For example, to extend the previous example to include a pairing
from `PDL1+ CK+` to `CD68+` cells, where `CD68+` is an existing phenotype,
extend the `pairs` argument without changing `phenotype_rules`:


```{r}
pairs <- list(
  c('PDL1+ CK+', 'T Cell'),
  c('PDL1+ CK+', 'CD68+'))
phenotype_rules <- list(
  'PDL1+ CK+'=list('CK+', ~`Entire Cell PDL1 (Opal 520) Mean`>3),
  'T Cell'=c('CD8+', 'FoxP3+')
)
```
