---
title: "Phenotype Spatial Distribution"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_vignette
---

```{r parameters, echo=FALSE,include=FALSE,message=FALSE}
# This document creates a spatial distribution report for two phenotypes in a
# single frame with optional positivity thresholds for each phenotype. It
# requires a cell seg table with phenotypes Optionally, a tissue segmentation
# TIFF image, if provided it will be used as a background for the image plots.
library(ggplot2)
library(magrittr)
library(spatstat)
```

<style type="text/css">
code, pre { 
  color: black;
  border: none;
  background-color: white;
  padding: 0;
  font-size: 14px;
}
.table {
    width: auto !important;
    max-width: 100%;
    margin-bottom: 20px;
}
</style>

```{r setup, echo=FALSE,include=FALSE,message=FALSE}
knitr::opts_chunk$set(echo=FALSE,fig.width=11, fig.height=8, 
                      comment=NA, warning=FALSE, message=FALSE)

data = read_cell_seg_data(cell_seg_path)

pheno1 = list(
  name=phenotypes[1], color=colors[1])

pheno2 = list(
  name=phenotypes[2], color=colors[2])

# Get point pattern datasets and distance matrices; direction matters...
window = spatstat::owin(xrange=c(0, max(data$`Cell X Position`)),
                        yrange=c(0, max(data$`Cell Y Position`)))
pheno_data1 = phenotype_as_ppp(data, pheno1, window)
pheno_data2 = phenotype_as_ppp(data, pheno2, window)

nn_dist12 = find_nearest_neighbor(pheno_data1, pheno_data2)
nn_dist21 = find_nearest_neighbor(pheno_data2, pheno_data1)

# Combined data for some reports
# Both phenotypes in one ppp
pp = spatstat::superimpose(pheno_data1$pp, pheno_data2$pp)

# Segmented image for background
if (grepl('cell_seg_data.txt', cell_seg_path))
{
  image_endings = c(
    'image_with_tissue_seg.tif',
    'image_with_tissue_seg.jpg',
    'tissue_seg_map.tif',
    'tissue_seg_map.jpg',
    'composite_image.tif',
    'composite_image.jpg'
  )
  for (ending in image_endings) {
    tissue_seg_path = sub('cell_seg_data.txt', ending, cell_seg_path)
    if (file.exists(tissue_seg_path)) break
  }
}

if (exists('tissue_seg_path') && file.exists(tissue_seg_path))
{
  if (grepl('jpg$', tissue_seg_path))
    background = jpeg::readJPEG(tissue_seg_path)
  else background = tiff::readTIFF(tissue_seg_path)
  
  # Tone it down a bit by bringing it closer to neutral gray
  background = (background - 0.75) / 8 + 0.75 
  background = as.raster(background)
  xlim = dim(background)[2] / pixels_per_micron
  ylim = dim(background)[1] / pixels_per_micron
} else {
  # If no background image guess at dimensions
  background = NA
  xlim = guessXMax(data)
  ylim = guessYMax(data)
}

# Remove def'n of tissue_seg_path so we will recompute next time for new image
rm(tissue_seg_path)

# Color palette for plots
palette = c(pheno1$color, pheno2$color)
names(palette) = c(pheno1$name, pheno2$name)
legend=TRUE
```

### Data
Spatial distribution of phenotypes from

``r cell_seg_path``

#### Selected phenotypes  

```{r results='asis'}
cat('-', pheno1$name, '\n')
cat('-', pheno2$name, '\n')
```


### First-order statistics

```{r results='asis'}
cat('Number of cells:', nrow(data), '\n\n')

data %>% dplyr::group_by(Phenotype) %>% 
  dplyr::summarize(Count=n(), 
                   Proportion=scales::percent(round(Count/nrow(data), 2))) %>% 
  knitr::kable(caption='Cell count and fraction of total per phenotype',
               table.attr='style="width: 30%"')

```

### Cell and phenotype locations

#### All cells  

```{r}
p = ggplot(data, aes(x=`Cell X Position`, y=`Cell Y Position`, color=Phenotype))
p = add_scales_and_background(p, background, xlim, ylim)
p = p + geom_point()
p + labs(title='Locations of all cells')
```

```{r fig.height=6+2*(length(levels(data$Phenotype))-1) %/% 3}
p + facet_wrap(~Phenotype) + guides(color=FALSE)
```

#### Selected phenotypes only  

```{r}
p = ggplot(as.data.frame(pp), aes(x=x, y=y, color=marks))
p = add_scales_and_background(p, background, xlim, ylim)
p = p + geom_point()
p = p + scale_color_manual('Phenotype', values=palette)
p = p + labs(x='Cell X Position', y='Cell Y Position',
             title='Locations of selected phenotypes')
p
```

```{r fig.height=6}
# p from previous chunk...
p + facet_wrap(~marks) + guides(color=FALSE)
```

### Nearest neighbors, selected phenotypes

```{r}
nn_plot(pheno_data1, pheno_data2, nn_dist12, background, xlim, ylim)
nn_plot(pheno_data2, pheno_data1, nn_dist21, background, xlim, ylim)
```

### Spatial statistics

#### Second-order statistics for selected phenotypes - G function 

G shows the distribution of nearest neigbors. 
It is the cumulative distribution function of the distance from
a typical cell to its nearest neighbor.

```{r, fig.height=5}
# Show G in a grid, cross-G only
x_label_microns = expression(paste('radius (', mu, 'm)'))
g_correction = 'km' # Kaplan-Meier edge correction only
plot_G = function(g, main)
{
  plot(g, cbind(km, theo) ~ r, legend=legend, 
       xlim=c(0,plot_gx_max), ylim=c(0,1),
       xlab=x_label_microns, main=main)
  micron_frac = as.function(g)(micron_line)
  abline(h=micron_frac, col = "lightgray", lty = 3)
  abline(v=micron_line, col='lightgray', lty=3)
  text(0, micron_frac, round(micron_frac, 2), adj=c(0, 0.4))
}

if (length(levels(pp$marks))<=1)
{
  cat('Not every phenotype has cells.\n')
} else {
  oldpar=par(mfrow=c(1,2))
  ignore = plot_G(Gcross(pp, pheno1$name, pheno2$name, correction=g_correction),
       main=bquote('Nearest neighbor,' 
                   ~ italic(.(pheno1$name) ~ 'to' ~ italic(.(pheno2$name)))))
  
  ignore = plot_G(Gcross(pp, pheno2$name, pheno1$name, correction=g_correction),
       main=bquote('Nearest neighbor,' 
                   ~ italic(.(pheno2$name)) ~ 'to' ~ italic(.(pheno1$name))))
  
  par(oldpar)
}
```


#### Cross-correlation for selected phenotypes  

Cross-correlation summarizes the relative density of cells of one type from cells of a second type.

```{r, fig.height=5}
pcf_correction = 'Ripley'

if (length(levels(pp$marks))<=1)
{
  cat('Not every phenotype has cells.\n')
} else {
  oldpar=par(mfrow=c(1,2))
  ignore = plot(pcfcross(pp, pheno1$name, pheno2$name, correction=pcf_correction), 
                legend=legend, xlab=x_label_microns, 
       main=bquote('Cross-correlation from' ~ italic(.(pheno1$name)) ~ 'to' 
                   ~ italic(.(pheno2$name))))
  
  ignore = plot(pcfcross(pp, pheno2$name, pheno1$name, correction=pcf_correction), 
                legend=legend, xlab=x_label_microns, 
       main=bquote('Cross-correlation from' ~ italic(.(pheno2$name)) ~ 'to' 
                   ~ italic(.(pheno1$name))))
  
  par(oldpar)
}
```
